---
title: "Creating an R data package"
author: "Luna L Sanchez Reyes"
date: "2023-04-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(usethis)
```

## Creating a new repository locally as an R package

With RStudio:
1. Click option to create a new project
2. Create a New directory
3. Choose create an R package
4. Give a name to the package
5. Click on `create in new session`

Once we have a backbone for an R package, start customizing it
- Modifying the DESCRIPTION file
- This files contains the metadata of the package
Metadata is "data about data", it contains the information about a data set.

- License GPL (>=2) is the most common for data packages
- Say something about the particularities of a GPL license
- Our package will only depend on R itself for now, that is no other packages 
  - It can be the latest version of R or any version that we are using


### Add a README file

Use a new package that is called `usethis`

```{r, eval = FALSE}
install.packages("usethis")
library(usethis)
```
The function `use_readme_md()` create a template for a README file in markdown
```{r}
use_readme_md()
```





