---
title: "functions"
author: "Luna L Sanchez Reyes"
date: "2023-03-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There is a difference between creating a function and running it.

What is it?


## The general form of functions

```{r, error = TRUE}
name_of_the_function <- function(data_frame, mass, length){
  data_frame %>% 
    group_by(mass, length) %>% 
    summarize(n = n())
}
```


```{r}
convert_pounds_to_grams <- function(pounds) {
     grams = 453.6 * pounds
     return(grams)
}
convert_pounds_to_grams(pounds = 13)
```




## Exercise 1: Converting between units of mass

1. Copy the following function into an R chunk of your Rmd file and replace the spaces ________ with variables names for the input and output.

```
 convert_pounds_to_grams <- function(________) {
     grams = 453.6 * pounds
     return(________)
 }
```

```{r}
red <- 3.75
yellow <- "Luna"
pounds = 3
convert_pounds_to_grams <- function(pounds = 1) {
   grams = 453.6 * pounds
   return(grams)
}

convert_pounds_to_grams2 <- function(grams = 1) {
   grams = 453.6 * pounds
   return(grams)
}
```

```{r}
convert_pounds_to_grams(pounds = 10089)
convert_pounds_to_grams()
length(x = yellow)

convert_pounds_to_grams2()
convert_pounds_to_grams2(grams = 3)
```



2. Use the function to calculate how many grams there are in 3.75 pounds.
What are the following parts of the function called:
a)
  convert_pounds_to_grams
b)
  {
   grams = 453.6 * pounds
   return(________)
  }
c)
  function(________)
  
  
## Exercise 2: The weight of dinosaurs

```{r}
get_mass_from_length_theropoda <- function(length){
  mass <- 0.73 * length ^ 3.63
  return(mass)
}
```


1. Use the function get_mass_from_length_theropoda() to get the mass of a Spinosaurus that is 16 m long based on its reassembled skeleton.


```{r}
get_mass_from_length_theropoda(16)
```


2. Create a new version of this function called get_mass_from_length() that takes the arguments length, a and b. Remember to change the body of the function accordingly, and use the following code to estimate the mass: 

mass <- a * length ^ b.

```{r}
get_mass_from_length <- function(length, a, b){
    mass <- a * length ^ b
    return(mass)
}
```



3. Use this new function to estimate the mass of a Sauropoda dinosaur (with parameters a = 214.44, and b = 1.46) that is 26 m long.

```{r}
get_mass_from_length(length = 26, a = 214.44, b = 1.46)
```


## Why do we use functions in programming?

- We already use functions a lot. Basically everything we run in R is a function.
- Reduce the amount of code we write.
- No need to repeat code, so less chance of repeating errors or introducing new errors by miss-typing code.
- Helps with reproducibility: no need to repeat code, allows others (and future self) to run code in a straightforward way, with accuracy.
- Make it easier to write modular code that you can reuse for other projects
- Makes it easier to run code and remember it.
- It allows to organize analysis better, by putting code intended to run something together in the same function.


## Exercise 4: Combining Functions

1. Write a function called convert_kg_to_pounds that converts kilograms into pounds using the formula pounds = 2.205 * kg.

```{r}
convert_kg_to_pounds <- function(kg = 1) {
  pounds <- 2.205 * kg
  return(pounds)
}
```


2. Use the function to calculate how many pounds is 50 kg.

```{r}
convert_kg_to_pounds(kg = 50)
```


3. Combine the function convert_kg_to_pounds() and your get_mass_from_length() function from the previous exercise to estimate the weight, in pounds, of a 12 m long Stegosaurus with a = 10.95 and b = 2.64 (The estimated a and b values for Stegosauria from Seebacher 2001).

```{r}
get_mass_from_length(length = 12, a = 10.95, b = 2.64) %>% 
  convert_kg_to_pounds()


length_in_pounds <- function(length, a, b){
  get_mass_from_length(length = length, a = a, b = b) %>% 
  convert_kg_to_pounds()
}
```


```{r}
length_in_pounds(length = 35, a = 13, b = 1.3)
```

Exercise 5: A function for the Portal data set

2. Combine the surveys and species tables into a single data frame.

There are several ways to create data frames within R. Depending on the data that we want as data frame, there are different functions that we can use to create a data frame object.
- `read.csv()`: If we already have a file (either CSV or TSV) with data that we want to read into R as a data frame 
- `data.frame()`: If we want to create a data frame from vectors
- `_join()` function: If we already have some data frames in R that we want to combine.
- There are many functions that create a data frame as an output, for example:
  - summarize(): It creates a new data frame summarizing information from another data frame
  - but there are many other packages that also generate data frames with different types of data
  
```{r}
combined_table <- inner_join(surveys, species, by = "species_id")
```


3. Write a function that:
Takes three arguments - a data frame (the combined table created before), a genus name, and a species name,
Uses functions from the dplyr package to produce a data frame with two columns: year and count, where count is the number of individuals (i.e., the number of rows) for the species indicated by genus and species in that year,
The function returns the resulting data frame. Note that this data frame is a time series by year.

As opposed to data frames, there is one one way to create a function: the function `function()`:

```
name_of_the_function <- function(argument1, argument2, arguent3, ...){
  This is the body of the function
  In here we do operation using the values stored in the arguments.
}
```

```{r}
# finder <- 
# species_yearcount
# my_function1
# test
# time_series
# my_function
# get_species_count_by_year

my_function <- function(mixed)
my_function <- function(a)
my_function <- function(combined_df)
my_function <- function(data_frame, genus_name, species_name) {
  data_frame %>% 
    group_by(year) %>% 
    filter(genus == genus_name, species == species_name) %>% 
    summarize(count = n()) %>% 
    return()
}

```

4. Use your function to get the data frame of time series by year for the genus Dipodomys and the species merriami and then make a graph of this time series using ggplot that has year on the x axis, count on the y axis, and displays the data as points.

```{r}
my_function(data_frame = combined_table, genus_name = "Dipodomys", species_name = "merriami")
```


Use your function to get the data frame time series by year for the genus Chaetodipus and species penicillatus and then make a graph of this time-series using ggplot that has year on the x axis, count on the y axis, and displays the data as blue points (with size = 1) connected by blue lines (with size = 2). Change the x axis label to Year and the y axis label to Number of Individuals.




