---
title: "if-else conditions"
author: "Luna L Sanchez Reyes"
date: "2023-03-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Logical and conditional statements: Review

These are pieces of code that return the `TRUE` or `FALSE` values, that is, a logical value.

The common operators of logical statements are:

- equality `==`
- inequality `!=`
- greater than `>`
- less than `<`
- greater or equal than `>=`
- less or equal than `<=`

The conditional statements allow to test several logical conditions at a time.
The condition operators (or symbols) are:

- AND `&` (inside dplyr functions we can also represent AND using a `,`)
- OR `|`

We also have logical functions that test if something is `TRUE` or `FALSE`, for example:

- `is.na()` is a function that tests if a value is an `NA`
- This function is part of a whole family of functions, they all start with `is.`:
- `is.vector()`
- `is.data.frame()`
- `is.factor()`

**_For next class: how to get all functions from a family (method)._**

- `which()` : takes logical vectors, it will give you the numerical index (position) of all values that are `TRUE`

```{r}
letters == "r"
which(letters == "r")
letters[18]
```

### Exercise 6

Create the following variables:

```{r}
 w <- 10.2
 x <- 1.3
 y <- 2.8
 z <- 17.5
 colors <- c("red", "blue", "green")
 masses <- c(45.2, 36.1, 27.8, 81.6, 42.4)
 dna1 <- "attattaggaccaca"
 dna2 <- "attattaggaacaca"
```
Use them to print whether or not the following statements are TRUE or FALSE.

a) w is greater than 10
b) "green" is in colors
c) x is greater than y
d) Each value in masses is greater than 40.
e) 2 * x + 0.2 is equal to y
f) dna1 is the same as dna2
g) dna1 is not the same as dna2
h) w is greater than x, or y is greater than z
i) x times w is between 13.2 and 13.5

```{r}
x * w < 13.5
x * w > 13.2
```

```
13.2 < x * w < 13.5 
```
This is how we would do it on paper, but in R  we ca only compare things in pairs.
For this we use the conditional statements:

```{r}
x *w < 13.5 & x * w > 13.2
```

j) Each mass in masses is between 30 and 50.



## How to make simple choices with `if ()`

The general structure of an if statement:

```
if (condition is TRUE) {
  Run all lines
  of code in 
  this block
  of code
}
```

If the condition is not `TRUE`, then nothing happens.

### Exercise 7
Complete the following if statement so that if `age_class` is equal to `“sapling”` it sets `y <- 10`:

```
 age_class = "sapling"
 if (){

 }
 y
```

Create the variable `age_class`:
```{r}
age_class = "sapling"
```

How do you test if `age_class` is equal to "sapling"?
```{r}
age_class == "sapling"
```

Now, complete the `if` statement:

```{r}
if (age_class == "sapling") {
  y <- 10
}
y
```

Remember! Inside the parentheses you have to write a logical or conditional statement.
If you forget the double equal sign, R will think you are trying to create a variable, and will throw an error:

```{r error = TRUE}
if (age_class = "sapling") {
  y <- 10
}
y
```


## Case when we have two options: `if else` structure

The general form of this structure:

```
if (condition) {
  code that runs if condition IS met
} else {
 code that runs if condition is NOT met
}
```

### Exercise 8:

Copy the following code and complete the if statement so that if `age_class` is equal to `“sapling”` it sets `y <- 10` and if `age_class` is equal to `“seedling”` it sets `y <- 5`.

We can solve this in a couple different ways.

First test if the variable is equal to "sapling", then cover all other conditions within the `else` block:
```{r}
 age_class = "seedling"
 if (age_class == "sapling"){
  y <- 10
 } else {
   print(age_class == "seedling")
   y <- 5
 }
y
```

Or, test first if `age_class` is equal to "seedling", then cover anything else:
```{r}
if (age_class == "seedling") {
  y <- 5
} else {
  y <- 10
}
```

### Handle more than 2 choices (3 choices or more)

In this case we are using the `else if` structure:


```
if (condition1) {
 first block code that is execites if condition 1 is met
} else if (condition2) {
 second block code that executes if condition2 is met
} else if (condition3) {
 more code
} else {
  this will cover all the conditions that are not specified before
}
```

- You do not have to end up with and `else` block.
- `else if` blocks are more intentional with the conditions.
- A simple `else` will run in all other cases no matter what.


### Exercise 9

Complete the if statement so that if `age_class` is equal to “sapling” it sets y <- 10 and if `age_class` is equal to “seedling” it sets y <- 5 and if `age_class` is something else then it sets the value of y <- 0.

Start with `age_class = "adult"`.

```{r}
 age_class = "adult"
 if (age_class == "sapling"){
  y <- 10
 } else if (age_class == "seedling") {
   y <- 5
 } else {
   y <- 0
}
y
```


