---
title: "Joining Vectors"
author: "Luna L Sanchez Reyes"
date: "2023-03-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## What are vectors?

- they are unidimensional matrices
- they can only hold one type of data, either numeric (integer or double), character, or logical (complex numbers)

## What are data frames

- it is a two dimensional matrix: rows, and columns
- it can hold any type of data
- it can only hold different types of data in a certain way:
  - only columns can have different data types
  - within a column, all rows must have the same data type

- A data frame can also be defined as a collection of vectors (they can be of different or the same type) all of the same length!

```{r}
surveys <- read.csv(file = "../data-raw/surveys.csv")
str(surveys)
```
## Creating vectors - review

A character vector with the function `c()`
```{r}
c("luna", "Avi", "Anita", "James", "Charles", "Damian", "Davinder") -> our_names
str(our_names)
1:7 # the colon operator creates a vector of numbers
1:7 -> my_numbers
-100:200
```
In how many steps does the colon operator increase? It increases in a step of 1.

What do we do if I want to create a numeric sequence that increases in steps different than 1?
?seq
```{r}
seq(-100, 200, by = 0.1) -> my_numbers
str(my_numbers)
```
```{r}
letters
LETTERS
```


## Creating data frames from vectors

The most general way to do this is with the function `data.frame()`:

```{r error = TRUE}
data.frame(names = our_names, numbers = my_numbers)
```
Remember: Vectors must have the same length (or be. amultiple of each other) to be part of a data frame!

```{r}
data.frame(names = our_names, numbers = 1:7)
```

R will recycle the values only if they are a multiple of the vector:

```{r}
data.frame(names = our_names, numbers = 1)
```

To recylce the values of a numeric vector of length 2, we have to repeat the vector of names two times, so it is a multiple of 2:

```{r}
data.frame(names = rep(our_names, 2), numbers = c(2, 5.5))
```

## Exercise 6

You have data on the length, width, and height of 10 individuals of the yew Taxus baccata stored in the following vectors:

```{r}
my_length <- c(2.2, 2.1, 2.7, 3.0, 3.1, 2.5, 1.9, 1.1, 3.5, 2.9)
width <- c(1.3, 2.2, 1.5, 4.5, 3.1, NA, 1.8, 0.5, 2.0, 2.7)
height <- c(9.6, 7.6, 2.2, 1.5, 4.0, 3.0, 4.5, 2.3, 7.5, 3.2)
```
Make a data frame that contains these three vectors as columns along with a "genus" column containing the genus name Taxus on all rows and a "species" column containing the species epithet baccata on all rows.

```{r}
data.frame(Length = my_length, Width = width, Height = height, genus = "Taxus", species = "baccata")
```
?table
?length
```{r error = TRUE}
my_table <- data.frame(width, length = my_length, height)
my_table$genus <- "Taxus"
my_table$species <- "baccata"
str(my_table)
my_table
```
## Extracting/accessing values from vectors and data frames

```{r}
surevys <- read.csv(file = "../data-raw/surveys.csv")
str(surveys)
```

One common way to extract or access vectors crom column in a data frame is the dollar sign symbol `$`
```{r}
surveys$record_id -> record_id
```

Another way is with the square brackets `[]`

```{r}
surveys[1:10, "hindfoot_length"]
```
If I want all the values from the rows of column hindfoot length:

```{r}
surveys[ , "hindfoot_length"] -> hindfoot_length
```

Another way is using doubel square brackets
```{r}
surveys[["record_id"]] %>% 
  head()
```

Exercise 7

1. Use $ to extract the weight column into a vector called surveys_weight

```{r}
surveys$weight -> surveys_weight
surveys_weight <- surveys$weight
```


2. Use [] to extract the month column into a vector called surveys_month

```{r}
surveys[, "month"] -> surveys_month
# if you use single square brackets you will still get a data frame
surveys["month"] %>% head()
# so we need double square bracket to get the actual vector
surveys[["month"]] %>% head()
```


3. Extract the hindfoot_length column into a vector and calculate the mean hindfoot length ignoring missing values.

```{r}
mean(surveys$hindfoot_length, na.rm = TRUE)
```

```{r}
surveys$hindfoot_length -> hindfoot_length
na.omit(hindfoot_length) %>% mean()
```

```{r}
surveys[["hindfoot_length"]] %>% mean(na.rm = TRUE)
surveys[,"hindfoot_length"] %>% mean(na.rm = TRUE)
```






