---
title: "loop exercises"
author: "Luna L Sanchez Reyes"
date: "2023-04-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 6: Size Estimates By Name Loop

If dinosaur_lengths.csv is not already in your working directory download a copy of the data on dinosaur lengths with species names. Load it into R.

```{r}
dino_data <- read.csv(file = "../data-raw/dinosaur_lengths.csv")
head(dino_data)
```

Write a function mass_from_length() that uses the equation `mass <- a * length^b` to estimate the mass of a dinosaur from its length. This function should take two arguments, length and species. For each of the following inputs for species, use the given values of a and b for the calculation from Seebacher 2001:

For Stegosauria: a = 10.95 and b = 2.64
For Theropoda: a = 0.73 and b = 3.63
For Sauropoda: a = 214.44 and b = 1.46
For any other value of species: a = 25.37 and b = 2.49
```{r}
mass_from_length <- function(dino_length, dino_species) {
  if (dino_species == "Stegosauria") {
    a <- 10.95
    b <- 2.64
  } else if (dino_species == "Theropoda") {
    a <- 0.73
    b <- 3.63
  } else if (dino_species == "Sauropoda") {
    a <- 214.44
    b <- 1.46
  } else {
    a <- 25.37
    b <- 2.49
  }
  mass <- a * dino_length^b
  return(mass)
}
```

Use this function and a for loop to calculate the estimated mass for each dinosaur, store the masses in a vector, and after all of the calculations are complete show the first few items in the vector using head().

```{r}
# for (i in 1:length(dino_lengths$species))

# seq_along(dino_lengths$species)

my_results <- vector(mode = "numeric", length = nrow(dino_data))
for (i in 1:nrow(dino_data)) {
  mass_i <- mass_from_length(dino_length = dino_data$lengths[i], dino_species = dino_data$species[i])
  # print(mass_i)
  my_results[i] <- mass_i
}

head(my_results)
```


Add the results in the vector back to the original data frame.

```{r}
dino_data$masses <- my_results
head(dino_data)
```


Show the first few rows of the data frame using head().

Calculate the mean mass for each species using dplyr.


```{r}
library(dplyr)
dino_data %>% 
  group_by(species) %>% 
  summarise(mean_mass = mean(masses))

```

## Exercise 7

You have a satellite collars on a number of different individuals and want to be able to quickly look at all of their recent movements at once. The data is posted daily to a URL, as a zip file, and contains one csv file for each individual: http://www.datacarpentry.org/semester-biology/data/individual_collar_data.zip.

Start your solution by:

If individual_collar_data.zip is not already in your working directory download the zip file using download.file()

```{r}
download.file(url = "http://www.datacarpentry.org/semester-biology/data/individual_collar_data.zip", destfile = "../data-raw/individual_collar_data.zip")
```

Unzip it using unzip()

```{r}
unzip("../data-raw/individual_collar_data.zip", exdir = "../data-raw/")
?unzip
```

Obtain a list of all of the files with file names matching the pattern "collar-data-.*.txt" (using list.files())

```{r}
list.files(path = "../data-raw/", pattern = "collar-data-.*.txt")
```

Use a loop to load each of these files into R and make a line plot (using geom_path()) for each file with long on the x axis and lat on the y axis. Graphs, like other types of output, won’t display inside a loop unless you explicitly display them, so you need put your ggplot() command inside a print() statement. Include the name of the file in the graph as the graph title using labs().

Add code to the loop to calculate the minimum and maximum latitude in the file, and store these values, along with the name of the file, in a data frame. Show the data frame as output.