---
title: "Tree join and viz"
author: "Luna L Sanchez Reyes"
date: "2023-04-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BiocManager)
library(ggtree)
library(ape)
```

## The package `ggtree` is on Bioconductor

There are a coupel of databases for hosting biological software in R, the most famous one is CRAN, but another very reliable one is called BioConductor.

When we are installing packages from CRAN, we use the function install.packages().
To install packages from Bioconductor, we use a function form an R package that is called `BiocManager` with the code `install.packages("BiocManager")`

Once that this package is installed, we can use one of its functions that is called `install()` to install packages from Bioconductor directly: `install("ggtree")`
If you get the following error:
```
Error: package or namespace load failed for ‘ggtree’ in loadNamespace(j <- i[[1L]], c(lib.loc, .libPaths()), versionCheck = vI[[j]]):
 there is no package called ‘aplot’
```
You will need to install another package with `install.packages("aplot")`


Exercise 1
Download this phylogenetic tree of species from the Portal Project Teaching Database, by clicking on the link and saving it to your data-raw folder.

Open the file by clicking on its name on the Files tab of RStudio’s Plots pane. It should look like this:

Read the file using a function from the ape pakage

```{r}
portal_tree <- read.tree(file = "../data-raw/portal-tree.tre")
portal_tree
```

Use the main function of the package ggtree (it is also called ggtree()) to visualize portal-tree.tre:

```{r}
ggtree(portal_tree)
```
```{r fig.height = 10, fig.width=7}
plot.phylo(portal_tree, cex = 0.5)
```


### The structure of a tree in R

We have an object that we named `portal_tree`, how do we explore its structure? The `str()` function is a great option
```{r}
str(portal_tree)
```

What is the class and length of this object:
```{r}
class(portal_tree)
```

Homework:How would you define a `phylo` object class based on its structure?


## Joining a tree and a data table

Download the portal species taxonomy and read it

```{r}
taxonomy <- read.csv("../data-raw/portal-species-taxonomy.csv")
head(taxonomy)
```

What types of functions you would use to joining 2 tables?
Functions from the `dplyr` package allow us to do that, for example `inner_join()`

```{r}
library(dplyr)
surveys <- read.csv(file = "../data-raw/surveys.csv")
species <- read.csv(file = "../data-raw/species.csv")
colnames(surveys)
colnames(species)
# we can use a function that will match the vcalues of tw vectors ot tell us which ones are present in the two:
intersect(colnames(surveys), colnames(species))
joined <- inner_join(surveys, species, by = "species_id")
head(joined)
```

Apply the same concept to join a phylogeetic tree and a table; the only difference is that we can't do inner_join, we can only do a left join:

```{r error = TRUE}
left_join(portal_tree, taxonomy, by = "label")
```











