---
title: "Tree from Open Tree"
author: "Luna L Sanchez Reyes"
date: "2023-04-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ape) # Analysis, Phylogenetics and Evolution R package
# install.packages("rotl") # for installation
library(rotl) # an R interface for the Open Tree of Life
```

Insects - 1 million named species
Mammals - 5k named species
Flowering plants - 200k named species 

### Read a newick tree from the Open Tree of Life

The following function can only read trees in newick format. Make sure your file is a newick tree.
If you got your tree from the Phylesystem repo, you tree is going to be in json format and will not be read by the function.

Newick Tree downloaded from https://tree.opentreeoflife.org/opentree/opentree13.4@ott118027/Gymnophiona

```{r}
tree <- read.tree(file = "../data-raw/delsuc_2008.tre")
```

Now plot the tree:

```{r, fig.height= 7, fig.width=5}
plot(tree, cex = 0.5)
mtext(text = "Delsuc 2008 OpenTree tree")
```

We can also use the url of a tree to read a tree into R. We will provide th eURL in the argument `file =`.
In this case, we do not have to keep a copy of the file, we can just download it directly 
The url is “http://ape-package.ird.fr/APER/APER2/primfive.tre”:

```{r}
small_tree <- read.tree(file = "http://ape-package.ird.fr/APER/APER2/primfive.tre")
small_tree
```
The R structure of a phylogenetic tree

### Getting a tree for your species from the Open Tree of Life

The Open Tree of Life has tools that allow to match the names of a group or species to their unified taxonomy so that we can look for information on their databases.

If we are doing this in R, we are using the `tnrs_match_names()` from the package `rotl`; tnrs stands for taxonomic name resolution service:

```{r}
tnrs <- tnrs_match_names(names = "amphibians")
class(tnrs)
tnrs
```
What is the main difference between the information that we have in the coumn `search_string` and `unique_name`:
One is the search string and the other is the scientific taxonomic name from the Open Tree of Life unified Taxonomy (OTT).

We can use the unique number identifier from OTT to get information fro that taxon.

To get a tree we can use the function `tol_subtree()`

```{r}
tol_subtree(ott_id = tnrs$ott_id)
```

### Exercise

Get a subtree from a group; extract the OTT id using the function `tnrs_match_names()`.





