---
title: "Tree from Open Tree"
author: "Luna L Sanchez Reyes"
date: "2023-04-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ape) # Analysis, Phylogenetics and Evolution R package
# install.packages("rotl") # for installation
library(rotl) # an R interface for the Open Tree of Life
```

Insects - 1 million named species
Mammals - 5k named species
Flowering plants - 200k named species 

### Read a newick tree from the Open Tree of Life

The following function can only read trees in newick format. Make sure your file is a newick tree.
If you got your tree from the Phylesystem repo, you tree is going to be in json format and will not be read by the function.

Newick Tree downloaded from https://tree.opentreeoflife.org/opentree/opentree13.4@ott118027/Gymnophiona

```{r}
tree <- read.tree(file = "../data-raw/delsuc_2008.tre")
```

Now plot the tree:

```{r, fig.height= 7, fig.width=5}
plot(tree, cex = 0.5)
mtext(text = "Delsuc 2008 OpenTree tree")
```

We can also use the url of a tree to read a tree into R. We will provide th eURL in the argument `file =`.
In this case, we do not have to keep a copy of the file, we can just download it directly 
The url is “http://ape-package.ird.fr/APER/APER2/primfive.tre”:

```{r}
small_tree <- read.tree(file = "http://ape-package.ird.fr/APER/APER2/primfive.tre")
small_tree
```
The R structure of a phylogenetic tree

### Getting a tree for your species from the Open Tree of Life

The Open Tree of Life has tools that allow to match the names of a group or species to their unified taxonomy so that we can look for information on their databases.

If we are doing this in R, we are using the `tnrs_match_names()` from the package `rotl`; tnrs stands for taxonomic name resolution service:

```{r}
tnrs <- tnrs_match_names(names = "amphibians")
class(tnrs)
tnrs
```
What is the main difference between the information that we have in the coumn `search_string` and `unique_name`:
One is the search string and the other is the scientific taxonomic name from the Open Tree of Life unified Taxonomy (OTT).

We can use the unique number identifier from OTT to get information fro that taxon.

To get a tree we can use the function `tol_subtree()`

```{r}
tol_subtree(ott_id = tnrs$ott_id)
```

### Exercise

Get a subtree from a group using the function `tol_subtree()`; extract the OTT id using the function `tnrs_match_names()`.

```{r}
tnrs_dolphins <- tnrs_match_names(names = "dolphins")
tnrs_dolphins
```
The function above ran the TNRS service to match the common name that I provided to the Open Tree of Life Taxonomy (OTT) and get a unique numeric identifier called the OTT id.
For dolphins this is the number `r tnrs_dolphins$ott_id`. 
The next step is to use the OTT id in the function that gets a subtree from Open Tree:

```{r}
tol_subtree(ott_id = tnrs_dolphins$ott_id) -> dolphins_subtree
```

```{r}
str(dolphins_subtree)
class(dolphins_subtree)
```

```{r}
plot.phylo(x = dolphins_subtree)
```

```{r}
Ntip(dolphins_subtree)
```

### Getting a tree for a set of taxa

Sometime we have very lareg groups and we just want to get a small tree representing some lineages inside the group.

For example, we just want a tree of all orders in amphibians. We know the common names of these which are frogs, salamanders and cecilians.

We want to find this lineages in the Open Tree of Life; the steps to do this are:

1. Get the scientific names of our taxa

```{r}
c("anura", "urodela", "gymnophiona") -> amphibian_orders
```

2. Match the scientific names to the OTT to get a unique numeric identifier

```{r}
tnrs_match_names(names = amphibian_orders) -> amph_order_tnrs
amph_order_tnrs
```

We have two columns at the beginning, search string contains the names that we provided, and unique_name contains the most updated or official or most commonly used taxonomic names for the search taxon that we provided.

3. Next stp is to use the OTT ids that we got to extract the tree conta8ining those lineages.
If we have only one name or lineage, we can use `tol_subree()`.
If we have more than one names, we will use `tol_induced_subtree()`.

```{r}
tol_induced_subtree(ott_ids = amph_order_tnrs$ott_id) -> amp_order_tree
amp_order_tree
```

```{r}
plot.phylo(x = amp_order_tree)
```

### Getting a tree of all families within amphibians


Install the `datelife` R package from GitHub, with a function from teh package `remotes`:

```{r}
library(remotes) # or install.packages("remotes")
install_github(repo = "phylotastic/datelife")
```

Now we can use the `datelife` package, but we have to load it first
```{r}
library(datelife)
```

We are going to use a function that extracts all the names from a taxon that belong to a given named rank. The function is called `get_ott_children()`:

```{r}
#?get_ott_children
get_ott_children(ott_ids = tnrs$ott_id, ott_rank = "family") -> amphibian_families
amphibian_families
```

Use the ott ids for all families to get an induced subtree:

```{r}
tol_induced_subtree(ott_ids = amphibian_families$Amphibia$ott_id) -> amp_fam_subtree
amp_fam_subtree
```
```{r, fig.height= 20}
plot.phylo(amp_fam_subtree)
```

Exercise:

1. Extract the names of the taxa from your data set for the final project.
2. Match them to the OTT using TNRS with the function `tnrs_match_names()`
3. Use the OTT ids to get an induced subtree with the function `tol_induced_subtree()`. 

I'll use the CSV file of the diet of Canis lupus, that we got form here https://datadryad.org/stash/dataset/doi:10.5061%2Fdryad.zgmsbccgg, as an example

```{r}
read.csv(file = "../data-raw/scat_RRA.csv") -> scat
head(scat)
```

We do not really have access to the subspecies names of the wolves, only their packs.
So we can start by creating a tree for the preys.
First, extract the scientific names:

```{r}
library(dplyr)
scat$Prey %>% 
  unique() %>% 
  tnrs_match_names() -> prey_ott_tnrs # match the scientific names to the Open Tree of Life Taxonomy

```

```{r error=TRUE}
prey_ott_tnrs %>% 
  filter(!is.na(ott_id)) -> prey_ott_tnrs

tol_induced_subtree(ott_ids = prey_ott_tnrs$ott_id)
```



