---
title: "UHURU data set visualization"
author: "Luna L Sanchez Reyes"
date: "2023-02-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A tip  for working with Rmarkdown
The Working directory inside this Rmarkdown r chunk is the following:
```{r eval = TRUE}
getwd()
```

Note: remember that working directories in a project and an R chunk are not always the same!

## 1. Describing the working data set

UHURU data set

```{r eval = FALSE}
knitr::include_graphics("https://www.google.com/url?sa=i&url=https%3A%2F%2Fen.wikipedia.org%2Fwiki%2FVachellia&psig=AOvVaw28QPvz2xvS6ULO7dQ-KDSD&ust=1677282495186000&source=images&cd=vfe&ved=0CBAQjRxqFwoTCLiomqzqrP0CFQAAAAAdAAAAABAE")
```



## 2. Reading the data set

We are reading a data set in TSV format. This is a "tab separated values" file.
To read it into R we use the function `read.csv()`  with the `sep =` argument set to `"\t"` which represents a tab in computer language.

```{r eval = TRUE, results='hide'}
read.csv(file = "../data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt", sep = "\t")
```

Assign the data to a variable so we can work with it

```{r eval = TRUE}
acacia <- read.csv(file = "../data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt", sep = "\t")
```

## 3. Acessing elements of a `data.frame`

It is similar to what we do for vectors, but we have two dimensions

```{r}

numbers <- 1:10
numbers
numbers[3:6]
numbers[c(1,5,7,3)]

acacia[,6]
str(acacia)
```

All the following are different ways of accessing a column or variable inside a data frame

```{r}
acacia$SURVEY
acacia[, 1]
acacia[ , "SURVEY"]
```

### Replacing elements in a data.frame

First we locate the element that we want to replace
```{r}
numbers[5] <- 100
numbers
```

## 4. Quality Control check

Check that everything is the correct class and type
```{r}
head(acacia)
str(acacia)
class(acacia$HEIGHT)
is.numeric(acacia$HEIGHT)
acacia$HEIGHT
as.numeric(acacia$HEIGHT)
```

We identified that HEIGHT should be numeric and is insetad character:

```{r}
acacia$HEIGHT <- as.numeric(acacia$HEIGHT)
acacia$HEIGHT
```

Coercion in R computer language means that a value was forced to be a type, in this case, the word "Dead" cannot be trasnformed to a number, so it is assigned the NA value


```{r}
acacia <- read.csv(file = "../data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt",
         sep = "\t",
         na.strings = "dead")
is.numeric(acacia$HEIGHT)
head(acacia)
# View(acacia)
```

## 5. Plotting data with `ggplot2`

Regular way to plot in R
```{r}
plot(x = acacia$HEIGHT, y = acacia$AXIS1, main = "Acacia height vs axis1")
?legend
```

With ggplot, we create layers

```{r}
library(ggplot2)

ggplot(data = acacia, mapping = aes(x = HEIGHT, y = AXIS1, color = TREATMENT)) +
  geom_point(size = 5) +
  labs(x = "Tree height", y = "Tree axis 1", title = "Acacia tree size relationships") 

```

```{r}
colors()
```
