---
title: "Data Visualization Homework - Acacia Vs Trees"
author: "Luna L Sanchez Reyes"
date: "2023-03-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
# getwd()
```

# Exercise 3

### Read the data

```{r}
trees <- read_tsv(file = "../data-raw/TREE_SURVEYS.txt")
head(trees)
```

---

### Quality assurance

After visually inspecting the variables, it looks like height is character type and it should be numeric

```{r}
str(trees)
head(trees$HEIGHT)
trees$HEIGHT <- as.numeric(trees$HEIGHT)
head(trees)
```

---

Alternatively, we can force variables to be assigned a specific type when we read them:
```{r}
trees <- read_tsv("../data-raw/TREE_SURVEYS.txt",
               col_types = list(HEIGHT = col_double(),
                                AXIS_2 = col_double()))
```

```{r}
canopy_area <- trees$AXIS_1 * trees$AXIS_2
head(trees)
trees$canopy_area <- canopy_area
trees$area <- canopy_area
```

```{r}
numbers <- -10:10
letters[-4]
numbers[c(1:3,5:10)]
numbers[-4]
trees[1:3,-(5:10)]
trees
```

---

Create a subset of the `trees` data frame with just the `SURVEY`, `YEAR`, `SITE`, and `canopy_area` columns:

```{r}
trees_test <- subset(trees, select=c(SURVEY, YEAR, SITE, canopy_area))
str(trees_test)
trees_test <- data.frame(SURVEY = trees$SURVEY, 
                         YEAR = trees$YEAR, 
                         SITE = trees$SITE, 
                         canopy_area = trees$canopy_area)
str(trees_test)
trees2 <- trees[,c("SURVEY", "YEAR", "SITE", "canopy_area")]
str(trees2)
```

---

Make a scatter plot with `canopy_area` on the x axis and `HEIGHT` on the y axis. Color the points by `TREATMENT` and create a subplot per species uding the function `facet_wrap()`. This will plot the points for each variable in the `SPECIES` column in a separate subplot. Label the x axis `“Canopy Area (m)”` and the y axis `“Height (m)”`. Make the point size 2.

```{r}
ggplot(data = trees, mapping = aes(x = canopy_area, y = HEIGHT, color = TREATMENT)) +
  geom_point() +
  labs(x = "Canopy area") +
  facet_wrap(~SPECIES) 
```

---

That’s a big outlier in the plot from (2). 50 by 50 meters is a little too big for a real acacia tree, so filter the data to remove any values for AXIS_1 and AXIS_2 that are over 20 and update the data frame. Then, remake the graph.

```{r}
# we can do this with the subset function
trees3 <- subset(trees, canopy_area <= 400)
# which(is.na(trees4$TREATMENT))

# OR finding the actual value that is the outlier
which(trees$canopy_area > 200)
trees[3839,]
trees4 <- trees[-3839,]
# n <- which(is.na(trees4$TREATMENT))
# trees4[n,]
```

The two tables are the same:

```{r}
ggplot(data = trees3, mapping = aes(x = canopy_area, y = HEIGHT, color = TREATMENT)) +
  geom_point() +
  facet_wrap(~SPECIES)
```

```{r}
ggplot(data = trees4, mapping = aes(x = canopy_area, y = HEIGHT, color = TREATMENT)) +
  geom_point() +
  facet_wrap(~SPECIES)
```

---

Final plot:

```{r}
ggplot(data = trees3, mapping = aes(x = canopy_area, y = HEIGHT, color = TREATMENT)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~SPECIES) +
  geom_smooth() +
  labs(x = "Canopy area", y = "Tree height")
```


Exercise 4

We want to compare the circumference to height relationship in acacia to the same relationship for all trees in the region. These data are stored in two different tables. 

One data set is in the file "ACACIA_DREPANOLOBIUM_SURVEY.txt"
The other data set is in "TREE_SURVEYS.txt".

First we load both data sets:

```{r}
# make sure to provide file name as relative path
read.csv(file = "../data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt",
         sep = "\t",
         na.strings = "dead") -> acacia


trees <- read_tsv("../data-raw/TREE_SURVEYS.txt",
               col_types = list(HEIGHT = col_double(),
                                AXIS_2 = col_double()))
```

Visual quality assurance and control with the function str()

```{r}
str(acacia)
is.numeric(acacia$CIRC)
is.numeric(acacia$HEIGHT)
```

```{r}
str(trees)
is.numeric(trees$CIRC)
is.numeric(trees$HEIGHT)
```

Make a graph with the relationship between CIRC and HEIGHT for all trees as gray circles in the background and the same relationship for acacia as red circles plotted on top of the gray circles. 

```{r}
ggplot() +
  geom_point(data = trees, mapping = aes(x = CIRC, y = HEIGHT), color = "gray", alpha = 0.5) +
  geom_point(data = acacia, mapping = aes(x = CIRC, y = HEIGHT), color = "red", alpha = 0.6)
```



```{r}
ggplot(data = trees, mapping = aes(x = CIRC, y = HEIGHT), color = "gray", alpha = 0.5) +
  geom_point(color = "gray") +
  geom_point(data = acacia, mapping = aes(x = CIRC, y = HEIGHT), color = "red", alpha = 0.6) +
  geom_smooth() +
geom_smooth(data = acacia, mapping = aes(x = CIRC, y = HEIGHT))
```


Scale both axes logarithmically. Include a linear model fitting for both sets of data, trying different linear models specified using the argument method =. Provide clear labels for the axes.

```{r}
?geom_smooth
ggplot(data = trees, mapping = aes(x = CIRC, y = HEIGHT), color = "gray", alpha = 0.5) +
  geom_point(color = "gray") +
  geom_point(data = acacia, mapping = aes(x = CIRC, y = HEIGHT), color = "red", alpha = 0.6) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm") +
geom_smooth(data = acacia, mapping = aes(x = CIRC, y = HEIGHT), method = "lm")
```


```{r}
str(trees)
unique(trees$SPECIES)
acacia_sub <- subset(trees, SPECIES == "Acacia_etbaica" | SPECIES == "Acacia_mellifera" | SPECIES == "Acacia_brevispica" | SPECIES == "Acacia_drepanolobium")
str(acacia_sub)
```

```{r}
other_sub <- subset(trees, SPECIES != "Acacia_etbaica" & SPECIES != "Acacia_mellifera" & SPECIES != "Acacia_brevispica" & SPECIES != "Acacia_drepanolobium")
str(other_sub)

other_sub <- subset(trees, SPECIES == "Croton" | SPECIES == "Balanites")
```

