---
title: "Portal data set"
author: "Luna L Sanchez Reyes"
date: "2023-03-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("dplyr")
library(dplyr)
```

## 1. Intro to the Portal data set

Homework: create an intro describing the location of the experiment and the different experimental treatments.
Paper here https://esajournals.onlinelibrary.wiley.com/doi/full/10.1890/15-2115.1

## 2. Load the data sets and create three data frames

```{r}
surveys <- read.csv(file = "../data-raw/surveys.csv")
head(surveys)
species <- read.csv(file = "../data-raw/species.csv")
head(species)
plots <- read.csv(file = "../data-raw/plots.csv")
head(plots)
```

------

## 3. The `dplyr` package for data wrangling 

### Subset columns from a data.frame wth the funvtion `select()`

```{r}
head(surveys)
surveys_subset <- select(surveys, month, day, year)

```

------

### Create new variables from existing variables or transform existing variables with `mutate()`

The hindfoot_length variable is measured in mm. I want a new variable that stores hindfoot length in cm.

```{r}
mutate(head(surveys), hindfoot_length_cm = hindfoot_length/10)
```


```{r}
head(surveys)
```

```{r}
surveys_mutated <- mutate(surveys, hindfoot_length_cm = hindfoot_length/10)
head(surveys_mutated)
```

------

### Sorting or ordering data with the function `arrange()`

If we want to order the data frame values based on the weight variable:

```{r}
surveys_arranged <- arrange(surveys, weight)
head(surveys_arranged)
```
Order values in descendant order with the function `desc()`

```{r}
surveys_arranged <- arrange(surveys, desc(weight))
head(surveys_arranged)
```

------

### Filter values with the function `filter()`

Filter the data frame to keep rows with weight values that are equal to 4:

```{r}
filter(surveys, weight == 4)
surveys_filtered <- filter(surveys, weight != 4)
head(surveys_filtered)
surveys_filtered <- filter(surveys, weight > 200)
head(surveys_filtered)
```

------

### Filter with more complex conditions

I want values that have weight larger than 200 AND also are females

```{r}
surveys_filtered <- filter(surveys, weight > 200, sex == "F")
head(surveys_filtered)
surveys_filtered <- filter(surveys, weight > 200 & sex == "F")
head(surveys_filtered)
```

```{r, results='hide'}
surveys_filtered <- filter(surveys, weight > 200 | sex == "F")
head(surveys_filtered)
```

------

### Filtering `NA` values

`NA` is a special valie in R. We can't use logical statements with it, we have to use the `is.na()` function:

```{r}
surveys_filtered <- filter(surveys, !is.na(weight))
head(surveys_filtered)
```




